### 2015-05-15 intro.tex
fixes
fix

Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software to provide an interface between the operating system and hardware devices that are attached to a computer. Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}. This has a negative impact in hardware product releases, as time-to-market is commonly dominated by driver development, verification and validation schedules~\cite{yavatkar2012era}.
%
Even after a driver has shipped, it is typically prone to many serious errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel; Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers. Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} found that they account for 19\% of the total bugs in Linux drivers, showcasing their significance. The majority of these concurrency bugs were found to be \emph{data races} or \emph{deadlocks} in various configuration functions and hot-plugging handlers.

@@

We present \whoop, an automated approach for static data race analysis in device drivers. \whoop is empowered by \emph{symbolic pairwise lockset analysis}, which attempts to prove a driver race-free by: (i) deriving a sound \emph{sequential} program that \emph{over-approximates} the originally concurrent program; (ii) instrumenting it to record \emph{locksets}; and (iii) using the locksets to assert that all accesses to the same shared resource are consistently protected by a common lock. Reducing analysis to reasoning over a sequential program avoids the need to enumerate thread interleavings, and allows reuse of existing successful sequential verification techniques.
%
We show that we can exploit the race-freedom guarantees provided by our symbolic analysis to achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a precise bug-finder used by Microsoft to analyze drivers and other concurrent programs. Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution.  By combining \whoop and \corral we achieve analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.  For two drivers, we observed even greater speedups of 12$\times$ and 20$\times$.
%
\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems as well as concurrent systems that use a similar programming model (e.g.\ file systems).

>>>>

Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software to provide an interface between the operating system and hardware devices that are attached to a computer. Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}. This has a negative impact on hardware product releases, as time-to-market is commonly dominated by driver development, verification and validation schedules~\cite{yavatkar2012era}.
%
Even after a driver has shipped, it typically has many undetected errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel; Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers. Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} found that they account for 19\% of the total bugs in Linux drivers, showcasing their significance. The majority of these concurrency bugs were found to be \emph{data races} or \emph{deadlocks} in various configuration functions and hot-plugging handlers.

@@

We present \whoop, an automated approach for static data race analysis in device drivers. \whoop is empowered by \emph{symbolic pairwise lockset analysis}, which attempts to prove a driver race-free by: (i) deriving a sound \emph{sequential} program that \emph{over-approximates} the originally concurrent program; (ii) instrumenting it to record \emph{locksets}; and (iii) using the locksets to assert that all accesses to the same shared resource are consistently protected by a common lock. Reducing analysis to reasoning over a sequential program avoids the need to enumerate thread interleavings, and allows reuse of existing successful sequential verification techniques.
%
We show that we can exploit the race-freedom guarantees provided by our symbolic analysis to achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a precise bug-finder used by Microsoft to analyze drivers and other concurrent programs. Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the Linux 4.0 kernel.  By combining \whoop and \corral we achieve analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.  For two drivers, we observed even greater speedups of 12$\times$ and 20$\times$.
%
\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems as well as concurrent systems that use a similar programming model (e.g.\ file systems).
